---
title: "ENVS-193DS Final"
author: "Sara Guefroudj"
date: 6/11/2025
format:
  html:
    toc: true # This line enables the Table of Contents
    toc-depth: 3 
---
[View this Github Repo](https://github.com/Saraguef123/ENVS-193DS_spring-2025_final)

# Final Set Up

```{r setup, warning = FALSE, message=FALSE}

 library(tidyverse)
library(janitor)
library(gt)
library(effsize)
library(here)
library(readxl)
library(scales)
library(ggeffects)
library(MuMIn)
library(dplyr)
library(RColorBrewer)
library(DHARMa)
sst <- read.csv(here("data","SST_update2023.csv"))
nest_boxes <- read.csv(here("data","occdist.csv"))

```

# Problem 1: Research Writing

## a.) Transparent statistical methods
In part 1 they used a person's R test because they specifically  state no correlation indicating that they are seeing if there is a correlation between the two variables being annual total nitrogen load and headwater distance. In addition they use numerical variables which indicate it's a persons test and not a spearman's test. In part 2 they use a one- way Anova test because the test describes how they are comparing means of different sources being urban land, fertilizer, and wastewater treatment across the one factor which is nitrogen load(kg year-1)

## b.) More information needed
One additional part  in part 2 the author could have addressed is a Post-HOC: Tukey HSD test because it would be meaningful to know which sources are different from each other. For example is there a difference in average nitrogen load between fertilizers and urban land.
Another addition they could have mentioned is the effective size because it would be more descriptive to know the magnitude of the differences between sources, not just saying they are significant with a p value because there could be a significant p value but we want to  know the magnitude and what it means in the context of the problem.

## c.)
We found a (size difference (η2= effect size) difference in average nitrogen load between sources(one-way ANOVA, F(?, ?) = value, p < 0.02, α = significance level).


# Problem 2. Data visualization 

##a.) Cleaning and Summarizing 
```{r}
sst_clean <- sst |> 
  clean_names() |> 
  select(temp,date) |> 
  mutate(
    date = mdy(date),
    year = year(date),
    month = month(date, label= TRUE, abbr = TRUE),
    month = as_factor(month),
    month = fct_relevel(
month,
"Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  ) |> 
filter(year %in% c(2018,2019,2020,2021,2022,2023)) |> 
  mutate(year = as_factor(year)) |> 
group_by(year,month) |> 
  summarize(
    mean_monthly_sst = round(mean(temp), 1),
    .groups = "drop")
sst_clean |> 
  ungroup() |> 
  select(year,month, mean_monthly_sst) |> 
  slice_sample(n = 5)
```

```{r}
str(sst_clean)
```

## b. Visualize the data 
```{r}
ggplot(data = sst_clean,
       aes(x=month,
           y= mean_monthly_sst,
           color = year,
           group = year))+
geom_point() +
  geom_line()+

labs(
  x = "Month",
  y = "Mean monthly sea surface temperature(°C)",
  color = "Year"
)  
  #scale_color_manual(values = )+
 #theme_classic()
  #theme(
#panel.grid.major = element_blank(),
#panel.grid.minor = element_blank(),
#panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
#legend.position = c(0.1,0.7))
#makes a gray border
    
  

```

# Problem 3. Data analysis

## a. Response variable
The 1s and 0s in this data describes if the nest boxes were occupied by each species with 0 being no and 1 being yes.

## b. Purpose of study
Swift parrots are an endangered species and the nest boxes were intended for them however they have different breeding patterns that lead to them not nesting in same location. Non target species are the common sterling which are native to the study area and can exploit the nest boxes. The tree martins however are nest competitors with both the common sterling and swift parrot

